State of the Practice for Medical Imaging Software

### 2021-11-18 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Udpates to SOP for MI up to Section 3.2 (List of Representative Software)

\usepackage{hyperref}
%\usepackage{tikz}
%\usetikzlibrary{mindmap}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{arydshln}

@@

We designed a general methodology to assess the state of the practice for SC
software \citep{SmithEtAl2021}. Details can be found in
Section~\ref{ch_methods}.  This methodology builds off prior work to assess the
state of the practice for such domains as Geographic Information Systems
\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
software \citep{Smith2018Seismology}, and Statistical software for psychology
\citep{smith2018statistical}.  In keeping with the previous methodology, we have
maintained the constraint that the work load for measuring a given domain should
be feasible for a team as small as one person, and for a short time, ideally
around a person month of effort. A person month is considered to be $20$ working
days ($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours
per day, or $20 \times 8 = 160$ person hours.

@@

We designed a general process for evaluating the state of the practice of
domain-specific SC software, that we instantiate for a specific scientific
domain.

Our method involves four steps:
\begin{enumerate}
\item choosing a software domain (Section \ref{sec_domain_selection});
\item collecting and filtering software packages (Section
\ref{sec_software_selection});
\item grading the selected software (Section \ref{sec_grading_software});
\item interviewing development teams for further information (Section
\ref{sec_interview_methods}).
\end{enumerate}

\noindent Section \ref{sec_applying_method} presents an example of how we
applied the method on the MI domain.

\subsection{Domain Selection} \label{sec_domain_selection}

Our methods are generic for any SC software, but they need to be applied to a
specific domain. When choosing a candidate domain, we prefer one with a large
number of active OSS. The reason is that we aim to finalize a list of 30
software packages \citep{SmithEtAl2021} after the screening step. For example, we
remove the ones without recent updates or specific functions. Besides, we prefer
OSS projects because our grading method requires access to the code. In
addition, we prefer a domain with an active community developing and using the
software. As a result, it is easier to invite enough developers for interviews.

We prefer 30 software packages providing similar functions or falling into
different sub-groups depending on our research purpose. So the domain needs to
have enough candidates in one sub-group or enough sub-groups to cross-compare.

We also prefer domains in which our team has expertise. We invite domain experts
to join and support our projects. They help us in many aspects, such as vetting
the software list and interview questions.

\subsection{Software Product Selection} \label{sec_software_selection}

The process of selecting software packages contains two steps: i) identify
software candidates in the chosen domain, ii) filter the list according to our
needs \citep{SmithEtAl2021}.

\subsubsection{Identify Software Candidates} \label{sec_identify_software_candidates}

We start with finding candidate software in publications of the domain. Then, we
search various websites, such as \hyperlink{https://github.com/}{GitHub},
\hyperlink{https://swmath.org/}{swMATH} and the Google search results for
software recommendation articles. We should also include the ones suggested by
the domain experts \citep{SmithEtAl2021}.

\subsubsection{Filter the Software List} \label{sec_filter_software_list}

The goal is to build a software list with a length of about 30
\citep{SmithEtAl2021}.

The only ``mandatory'' requirement is that the software must be OSS, as defined
in Section \ref{sec_open_source_software}. We need this because evaluating some
software qualities requires the source code.

The other filters are optional, and we consider them according to the number of
software candidates and the objectives of the research project. We apply them in
the following priority order:

\begin{enumerate}
\item The functions and purpose of the software. An SC domain often contains
software with various functions and purposes. For example, some MI software
packages are Tool Kit for developers to use, and some others offer Visualization
function to end-users. We have two options:
\begin{itemize}
\item selecting a set of software with the same major function. In this case, we
can use the identical process to assess all packages, e.g., same input to
measure \textit{Robustness}. Also, when we give impression scores to qualities
such as \textit{Installability} and \textit{Usability}, the results are more
comparable. Thus, it is more feasible to collect the results and rank the
software.
\item selecting software from a set of different sub-groups. For example, we can
choose 10 MI software from each of the three sub-groups: Visualization, Tool
Kit, and PACS. The downside: we may need different processes to measure each
sub-group; it may be less accurate to mix all three sub-groups and rank them
together. The benefit: we can cross-compare the development processes between
the sub-groups.
\end{itemize}

\item The version control tool. The empirical measurement tools listed in
Section \ref{sec_empirical_measurements} only work on projects using
\hyperlink{https://git-scm.com/}{Git}, so we prefer software with Git. Some
manual steps in empirical measurement depend on a few metrics of GitHub, which
makes projects held on GitHub more favored \citep{SmithEtAl2021}.

\item The age of the software. Some of the OSS projects may experience a lack of
recent maintenance. So we eliminate packages without recent updates, unless they
are still popular and highly recommended by the domain users
\citep{SmithEtAl2021}. We consider a software project as ``alive" if it has any
update within the last 18 months; otherwise, we mark it as ``dead".
\end{enumerate}

The order of filters 2 and 3 is flexible. We adjust it according to the number
of software packages affected by the filters, and the number of ones remaining
on the list.

\subsubsection{Vet the Software List} \label{sec_vet_software_list}

Before showing our filtered list to the domain experts, we ask them to list
their top 10 software in the domain. Then, we cross-compare the two lists and
discuss the commonality and variability.  In addition, we ask them to vet our
filtered list. They provide views on whether the list is reasonable. We also use
their opinions for a filtering process. For example, if a software package is
not OSS and has had no updates for a long time, but the domain experts identify
it as a valuable product, we still consider it in our final list.

@@

\subsubsection{Grading Template} \label{sec_grading_template}

@@

\subsubsection{Empirical Measurements} \label{sec_empirical_measurements}

@@

\subsubsection{Technical Details} \label{sec_technical_details}

To test the software on a ``clean'' system, we create a new virtual machine (VM)
for each software and only install the necessary dependencies before measuring.
We make all 30 VMs on the same computer, one at a time, and destroy them after
measuring.

@@

@@

\subsubsection{Interview Questions} \label{sec_interview_questions}

@@

\subsubsection{Interviewee Selection} \label{sec_interviewee_selection}

@@

\subsubsection{Interview Process} \label{sec_interview_process}

@@

\subsection{Applying the Method to MI} \label{sec_applying_method}

This section shows an overview of applying our method to the MI domain.

\subsubsection{Domain Selection}

Based on the principles in Section \ref{sec_domain_selection}, we selected the
MI domain and the sub-group of software with the Visualization function shown in
Figure \ref{fig_mi_functions}. We also included Dr. Michael Noseworthy, a
professor of Electrical and Computer Engineering at McMaster University,
Co-Director of the McMaster School of Biomedical Engineering, and Director of
Medical Imaging Physics and Engineering at St. Joseph’s Healthcare, and some of
his students as the MI domain experts in our team.

\subsubsection{Software Product Selection} \label{sec_mi_software_selection}

By using the method in Section \ref{sec_identify_software_candidates}, we
identified 48 MI software projects as the candidates from publications
\citep{Bjorn2017} \citep{Bruhschwein2019} \citep{Haak2015}, online articles related
to the domain \citep{Emms2019} \citep{Hasan2020} \citep{Mu2019}, forum discussions
related to the domain \citep{Samala2014}, etc. Appendix
\ref{ap_list_before_filtering} shows all 48 software packages.

Guided by the method in Section \ref{sec_filter_software_list}, we filtered the
list with a process as follows:

\item Among them, there were eight that we could not find their source code,
such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}. These packages
are likely to be freeware defined in Section \ref{sec_freeware} and not OSS. So
following guidelines in Section \ref{sec_filter_software_list} we removed them
from the list.

\item Next, we focused on the MI software providing Visualization functions, as
described in Section \ref{sec_scope}. Seven of the software on the list were
Tool Kits or libraries for other software to use as dependencies, but not for
end-users to view medical images, such as \textit{VTK}, \textit{ITK}, and
\textit{dcm4che}; another three were PACS. We also eliminated these from the
list.

\item Finally, we removed \textit{Open Dicom Viewer} from the list because it
had not received any updates for a long time (since 2011). After that, only
\textit{MatrixUser} and \textit{AMIDE} were considered as ``dead". However, both
of them had much more recent updates (after 2017) than \textit{Open Dicom
Viewer}.

We still preferred projects using git and GitHub and being updated recently, but
did not apply this filter since packages were already below 30. Even without
this filter, 27 out of the 29 software packages on the filtered list used git,
and 24 chose GitHub.

Following the process in Section \ref{sec_vet_software_list}, our domain experts
provided a list of top software that contains 12 software products (Table
\ref{tab_top_software_experts}). We compared two lists and found six common
ones.

\begin{table}[ht]
\centering
\begin{tabular}{ll}
\hline
Software & On both lists \\ \hline
3D Slicer & X \\
Horos & X \\
ImageJ & X \\
Fiji & X \\
AFNI &  \\
FSL &  \\
Freesurfer &  \\
Mricron & X \\
Mango & X \\
Tarquin &  \\
Diffusion Toolkit &  \\
MRItrix &  \\ \hline
\end{tabular}
\caption{\label{tab_top_software_experts}Top software by the MI domain experts}
\end{table}

We included \textit{Mango} in the initial list, but removed it because it was
not OSS. However, we kept \textit{Papaya}, a the web version of \textit{Mango}.
Instead of \textit{MRIcron}, we chose \textit{MRIcroGL}, because
\textit{MRIcron} development had moved to \textit{MRIcroGL} \citep{Rorden2021b}.

Six software packages on the domain experts' list were not on our filtered list.
We believed their primary function was Analysis mentioned in Section
\ref{sec_scope}. Thus, we did not include them in our final list.

After vetting our filtered list, the domain experts believed it was reasonable
and did not identify any problem. Thus, as shown in Appendix
\ref{ap_list_before_filtering}, eventually, we had 29 software products on the
final list.

\subsubsection{Grading Software}
\label{sec_applying_method_grading}
Then we followed the steps in Section \ref{sec_grading_software} to measure and
grade the software. 27 out of the 29 packages are compatible with two or three
different OS such as Windows, macOS, and Linux, and 5 of them are browser-based,
making them platform-independent. However, in the interest of time, we only
performed the measurements for each project by installing it on one of the
platforms, most likely Windows.

\subsubsection{Interviews}
\label{sec_apply_to_mi_interviews}
We received ethics clearance from the McMaster University Research Ethics Board
(Appendix \ref{ap_ethics}). Going through the interview process in Section
\ref{sec_interview_methods}, we contacted all of the 29 teams. Members from
eight teams responded and agreed to participate. As a result, we interviewed
nine developers and architects from the eight teams.

>>>>

\usepackage{hyperref}
%\usepackage{tikz}
%\usetikzlibrary{mindmap}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{arydshln}
\usepackage{paralist}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\mn}[1]{\authornote{magenta}{MN}{#1}} %Mike Noseworthy
\newcommand{\ad}[1]{\authornote{cyan}{AD}{#1}} %Ao Dong

\newcommand{\notdone}[1]{\textcolor{red}{#1}}
\newcommand{\done}[1]{\textcolor{black}{#1}}

@@

We designed a general methodology to assess the state of the practice for SC
software \citep{SmithEtAl2021}. Details can be found in Section~\ref{ch_methods}
and in \citep{Dong2021}.  This methodology builds off prior work to assess the
state of the practice for such domains as Geographic Information Systems
\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
software \citep{Smith2018Seismology}, and Statistical software for psychology
\citep{smith2018statistical}.  In keeping with the previous methodology, we have
maintained the constraint that the work load for measuring a given domain should
be feasible for a team as small as one person, and for a short time, ideally
around a person month of effort. A person month is considered to be $20$ working
days ($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours
per day, or $20 \times 8 = 160$ person hours.

@@

We developed a methodology for evaluating the state of the practice of SC
software \citep{SmithEtAl2021}.  The  methodology can be instantiated for a
specific domain of scientific software, which in the current case is medical
imaging software for visualization.  Our methodology involves and engages a
domain expert partner throughout, as discussed in
Section~\ref{sec_vet_software_list}.  The four main steps of the methodology are:

\begin{enumerate}
\item Identify list of representative software packages
(Section~\ref{sec_software_selection});
\item Measure (or grade) the selected software (Section~\ref{sec_grading_software});
\item Interview developers (Section~\ref{sec_interview_methods}).
\item Answer the research questions (as given in Section~\ref{sec_motivation})
\end{enumerate}

In the sections below we provide additional detail on the above steps, while
concurrently giving examples of how we applied the methodology to the MI domain.

\subsection{Interaction With Domain Expert} \label{sec_vet_software_list}

The Domain Expert is an important member of the state of the practice assessment
team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
software, or try to definitively rank the software. Non-experts have the problem
that they can only rely on information available on-line, which has the
following drawbacks:
\begin{inparaenum}[i)]
  \item the on-line resources could have false or inaccurate information; and,
  \item the on-line resources could leave out relevant information that is so
in-grained with experts that nobody thinks to explicitly record it.
\end{inparaenum}

Domain experts may be recruited from academia or industry.  The only
requirements are knowledge of the domain and a willingness to be engaged in the
assessment process.  The Domain Expert does not have to be a software developer,
but they should be a user of domain software.  Given that the domain experts are
likely to be busy people, the measurement process cannot put to much of a burden
on their time.  For the current assessment, our Domain Expert (and paper
co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
Engineering at McMaster University, Co-Director of the McMaster School of
Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
at St. Joseph’s Healthcare, Hamilton, Ontario, Canada.

In advance of the first meeting with the Domain Expert, the expert is asked to
create a list of top software packages in the domain.  This is done to help
the expert get in the right mind set in advance of the meeting.  Moreover,
by doing the exercise in advance, we avoid the potential pitfall of the expert
approving the discovered list of software without giving it adequate thought.

The Domain Experts are asked to vet the collected data and analysis.  In
particular, they are asked to vet the proposed list of software packages and the
AHP ranking.  These interactions can be done either electronically or with
in-person (or virtual) meetings.

\subsection{List of Representative Software} \label{sec_software_selection}

We have a two step process for selecting software packages: i) identify
software candidates in the chosen domain; and, ii) filter the list to remove less relevant members \citep{SmithEtAl2021}.

We initially identified 48 MI candidate software projects from the literature
\citep{Bjorn2017, Bruhschwein2019, Haak2015}, online articles \citep{Emms2019,
Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
48 packages is available in \citet{Dong2021}.  To reduce the length of the list
to a manageable number (29 in this case), we filtered the original list as
follows:

\item We removed the packages that did not have source code available, such as
\textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.

\item We focused on the MI software that provides visualization functions, as
described in Section~\ref{sec_scope}. We removed seven packages that were
toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}.
We removed another three that were for PACS.

\item We removed \textit{Open Dicom Viewer}, since it has not received any
updates for a long time (since 2011).

The Domain Expert provided a list of his top 12 software packages.  We compared
his list to our list of 29.  We found 6 packages were on both lists: \textit{3D
Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} (we
actually use the update version \textit{MRIcroGL}) and \textit{Mango} (we
actually use the web version \textit{Papaya}).  Six software packages
(\textit{AFNI}, \textit{FSL}, \textit{Freesurfer}, \textit{Tarquin},
\textit{Diffusion Toolkit}, and \textit{MRItrix}) were on the Domain Expert
list, but not on our filtered list.  However, when we examined those packages,
we found they were out of scope, since their primary function was not
visualization.  The Domain Expert agreed with our final choice of 29 packages.

@@

@@

@@

As in \citet{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
optimal testing environments for each package. VMs were used because it is
easier to start with a fresh environment without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors. When
filling in the measurement template spreadsheet, the the details for each VM
should be noted, including hypervisor and operating system version.

@@

Then we followed the steps in Section \ref{sec_grading_software} to measure and
grade the software. 27 out of the 29 packages are compatible with two or three
different OS such as Windows, macOS, and Linux, and 5 of them are browser-based,
making them platform-independent. However, in the interest of time, we only
performed the measurements for each project by installing it on one of the
platforms, most likely Windows.

@@

@@

@@

@@

We received ethics clearance from the McMaster University Research Ethics Board
(Appendix \ref{ap_ethics}). Going through the interview process in Section
\ref{sec_interview_methods}, we contacted all of the 29 teams. Members from
eight teams responded and agreed to participate. As a result, we interviewed
nine developers and architects from the eight teams.
